---
title: "project"
output: 
    md_document :
        toc: true
        variant: markdown_github
---

# Getting and Cleaning Data Course Project 

## run_analysis(pathToUciHarDataset)

## dependencies
```{r}
source("str_activity.R")
source("str_feature.R")
```

The variable ```pathToUciHarDataset``` contains the path to UCI HAR datasets
```{r}
pathToUciHarDataset = "UCI HAR Dataset"
```

## 1. Merges the training and the test sets to create one data set.

Load test data in ```test``` data frame
```{r, cache=TRUE}
test <- read.table(paste(pathToUciHarDataset,"/test/X_test.txt", sep=""))
```

Load train data in ```train```
```{r, cache=TRUE}
train <- read.table(paste(pathToUciHarDataset,"/train/X_train.txt", sep=""))
```

Merges all data in a single dataset named ```all_ds```. Check dimensions of the resulting dataset
```{r}
all_ds <- rbind(test, train)
dim(test)
dim(train)
dim(all_ds)
stopifnot (nrow(all_ds) == nrow(test) + nrow(train), ncol(all_ds) == ncol(test))
```

Remove non needed dataframes
```{r}
rm(train)
rm(test)
```

## 2. Extracts only the measurements on the mean and standard deviation for each measurement.

Load all features (measurements) in ```features``` dataframe. I used the csv variant, using as separtor white space, as all columns are separated evenly by white spaces
```{r}
features <- read.csv(paste(pathToUciHarDataset,"/features.txt", sep=""), sep=" ", header=F)
```

Check dimensions. ```features``` dataframe contains a row for each column in ```all_ds```
```{r}
dim(all_ds)
dim(features)
stopifnot(dim(all_ds)[2] == dim(features)[1])
```

Second column of ```features``` dataframe contains feature (column) name. We will follow approach such as any column containing -std() or -mean() is feature's standard deviation or mean, so this will become a selected (valid) feature for our work. The dataframe ```selected_features``` stores only selected features
```{r}
selected_features <- features[grepl("-mean()|-std()", features[,2]),]
```

Check ```selected_features```, so there are 79 features names which commits to the selection criteria
```{r}
dim(selected_features)
```

Extract only measurements from ```selected_features``` in ```selected_ds``` dataframe. Check resulting data frame dimensions
```{r}
selected_ds <- all_ds[,selected_features[,1]]
dim(selected_ds)
stopifnot(nrow(selected_ds) == nrow(all_ds), ncol(selected_ds) == nrow(selected_features))
```

Remove ```all_ds``` from workspace
```{r}
rm(all_ds)
```

## 3. Uses descriptive activity names to name the activities in the data set.

Load activity labels in ```activity_labels``` datframe.
```{r}
activity_labels <- read.csv(paste(pathToUciHarDataset,"/activity_labels.txt", sep=""), sep=" ", header=F)
```

Load test and training activities. I used the fwf variant, using width = 1 (only one digit).
Merge all activities together and check dimensions of resulting ```activities``` dataset. Remove non needed activities dataframes
```{r}
test_activities <- read.fwf(paste(pathToUciHarDataset,"/test/y_test.txt", sep=""), header=F,widths=1)
train_activities <- read.fwf(paste(pathToUciHarDataset,"/train/y_train.txt",sep=""), header=F,widths=1)
activities <- rbind(test_activities, train_activities)
dim(activities)
stopifnot(nrow(activities) == nrow(selected_ds))
rm(test_activities)
rm(train_activities)
```

Check dimensions of ```activities``` and validate min/max (10299 observations, min=1, max=6) aligned with activity labels first column
```{r}
dim(activities)
table(activities)
print (activity_labels)
```

Apply ```str_activity``` to activity labels and store result in new column. This step is not really needed, but I just wanted to check new factors were correct
```{r}
activity_labels$Activity = factor(activity_labels[,1], labels=str_activity(activity_labels[,2]))
str(activity_labels)
print(activity_labels)
```

Add new ```Activity``` column translating activity integer to its factor in the ```activities``` dataframe. Check translation against original form
```{r}
activities$Activity <- factor(activities[,1], labels = str_activity(activity_labels[,2]))
table(activities[,1])
table(activities$Activity)
```

Bind activity column with selected features. Check new dimensions (10299x80, same rows but plus one more column, which is the new activity label)
```{r}
selected_ds$Activity <- activities$Activity
dim(selected_ds)
stopifnot(ncol(selected_ds) == 80)
```


## 4. Appropriately labels the data set with descriptive variable names.

Check column names before translation. So we've got almost all column names following pattern V1:V552, except last column which is named Activity
```{r}
names(selected_ds)
```

Label columns with descriptive names, using function ```str_feature```. We used a little trick, the selected_features dataset populated before, adding the new column Activity we added in previously. Check columns
```{r}
names(selected_ds) <- c(str_feature(selected_features[,2]), "Activity")
names(selected_ds)
stopifnot(ncol(selected_ds) == nrow(selected_features) + 1)
str(selected_ds)
```

## 5. From the data set in step 4, creates a second, independent tidy data set with the average of each variable for each activity and each subject.

Load test and training subjects. Merge all subjects into ```subjects``` variable and check their dimensions. Remove non needed subject dataframes
```{r}
test_subjects <- read.fwf(paste(pathToUciHarDataset,"/test/subject_test.txt",sep=""), widths=2, header=F)
train_subjects <- read.fwf(paste(pathToUciHarDataset,"/train/subject_train.txt",sep=""), widths=2, header=F)
subjects <- rbind(test_subjects, train_subjects)
str(subjects)
stopifnot(nrow(subjects) == nrow(selected_ds))
rm(test_subjects)
rm(train_subjects)
```

Add column ```Subject``` to ```selected_ds``` dataframe. Check dimensions (same number of rows, but plus one more column = 81)
```{r}
selected_ds$Subject <- subjects[,1]
dim(selected_ds)
stopifnot(ncol(selected_ds) == 81)
```


Finally, we get the tidy dataset grouping by subject and activity name, and summarising the mean of the groups. Resulting dataset is clean up, removing last columns (original Activity and Subject columns)
```{r, warning=FALSE}
tidy_ds <- aggregate(selected_ds, by=list(Subject = selected_ds$Subject, Activity = selected_ds$Activity), mean)[,1:81]
```
```{r, results='asis'}
knitr::kable(tidy_ds)
```

## str_activity(x)
The variable ```x``` is the input data

Custom funciton which cleans up activity labels:

* to lower
* replaces "_" by white space

```{r}
str_activity <- function(x){ 
     gsub("_"," ",tolower(x))
}
```

## str_feature(x)
The variable ```x``` is the input data

Custom function which cleans up the feature names:

* remove "()"
* replace dimension name by "Dim{dimesion}""
* replace "-mean" by "Mean"
* replace "-std" by "Std"

This will help in case we need to manipulate columns in the feature using their names

```{r}
str_feature <- function(x){ 
    x <- gsub("[\\(\\)]", "", x)
    x <- gsub("-X$", "DimX", x)
    x <- gsub("-Y$", "DimY", x)
    x <- gsub("-Z$", "DimZ", x)
    x <- gsub("-mean", "Mean", x)
    gsub("-std", "Std", x)
}
```
